---
title: "R Notebook"
output: html_document
---

## Libraries
```{r}
library(dplyr)
library(MASS)
library(ggplot2)
library(reshape2)
library(caret)
library(ggpubr)
library(car)
```

```{r}
base_dir = "/Users/jlabelle/Dropbox (Partners HealthCare)/"
working_dir<- paste0(base_dir, "Personal/STAT_E109/Assignments/Project/")

## Set output directories
data_dir<- "analysis/Preprocessing/"
analysis_dir<- "analysis/predictCellTypeComp/"
figure_dir<- "figures/"

```

## Read in data:
1. Pseudobulked data used to build linear regression model. Each dataframe contains the proportion of that cell type found in each sample + expression of the top 15 genes contributing to that cell type

2. Bulk expression data- linear regression models applied to this data to predict cell type composition

```{r}
## Pseudobulk data
  ## List of files for each cell type
  files<- list.files(data_dir, pattern="_pseudobulked_marker.prop.csv")
  files<- files[!grepl("NK", files)]
  cell_types<- gsub("_pseudobulked_marker.prop.csv", "",gsub("-", "\\." ,files))
  
  ## Read in each df
  df_byCellType<- lapply(files, function(x){read.csv(paste0(data_dir, x), row.names = "X")})
  names(df_byCellType)<- cell_types
  
## Bulk data
bulk_cm<- readRDS(paste0(data_dir, "bulk_melanoma_cm_tpm_scaled.Rds"))

## Save head of data frame to show example of data structure
df<-df_byCellType$B.cells[1:5, ]
png(paste0(figure_dir, "Table2_Bcells_example.df.png"), height=200, width=2300)
p<-tableGrob(df)
grid.arrange(p)
dev.off()
```

## Split each dataframe into training/testing
```{r}
train.test_byCellType<- lapply(cell_types, function(x){
  df<- df_byCellType[[x]]
  set.seed(1)
  ind <- sample(2, nrow(df), replace = T, prob = c(0.7, 0.3)) 
  train <- df[ind == 1,]
  test <- df[ind == 2,]
  tt_list<- list(train=train, test=test)
  return(tt_list)
})
names(train.test_byCellType)<- cell_types
```

#######################
## Linear regression ##
#######################

## First, check to see if expression data is roughly normally distributed. 
In general, gene expression is assumed to be normally distributed- researchers regularly apply tests that assume normality to gene expression data. Representative genes are shown here, although not all genes show clear normal distribution.
```{r}
png(paste0(figure_dir, "Figure1_geneExpression_normDistribution.png"), width=600, height=300)
par(mfrow=c(2,3))
hist(df_byCellType$B.cells$CLEC17A, main="CLEC17A in B.cells")
hist(df_byCellType$CAFs$COL1A1,main="COL1A1 in CAFs")
hist(df_byCellType$Endothelialcells$MYCT1, main="MYCT1 in Endothelial cells")
hist(df_byCellType$Macrophages$C1QC,main="C1QC in Macrophages")
hist(df_byCellType$melanoma$PLP1,main="PLP1 in melanoma")
hist(df_byCellType$T.cells$CD3E, main="CD3E in T.cells")
dev.off()

```


## For each cell type, generate multivariate linear regression model from training data:
Outcome variable is the proportion of that cell type found in that sample
Predictor variables are normalized expression values for the top 15 genes contributing to that cell type
```{r}
## For each cell type, generate model using stepAIC to determine which of the 15 genes contribute 
all_models<- lapply(cell_types, function(cell_type){
  train<- train.test_byCellType[[cell_type]]$train
  train$Dependent<- train[[cell_type]]; train[[cell_type]]<-NULL
  model<- lm(Dependent ~ ., train)
  stepAIC_model<- stepAIC(model, trace=FALSE)
  return(stepAIC_model)
})
names(all_models)<-cell_types

## VIF for each model
all_vif<- lapply(all_models, function(x){
  tmp<- as.data.frame(vif(x))
  tmp$gene<- rownames(tmp); rownames(tmp)<-NULL
  return(tmp)
})
all_vif<- do.call("rbind", all_vif)

## Plot VIF values
colnames(all_vif)<- c("VIF", "Gene")
all_vif$CellType<-  gsub('[[:digit:]]+', '', rownames(all_vif))
ggplot(all_vif, aes(x=Gene, y=VIF, color=CellType))+
  geom_point()+
    theme_bw()+
  facet_grid(cols = vars(CellType), space="free_x", scale="free_x")+
  geom_hline(yintercept = 10, linetype="dashed")+
  theme(axis.text.x = element_text(angle = 45, hjust=1, face="bold", color="black"),
        legend.position = "none")+
  ylab("VIF") + xlab("gene input into model")
ggsave(paste0(figure_dir, "Figure2_LR_VIF_genesInModel.png"), width=8, height=3)

## Remove genes with high VIF (>10)
highVIF<- c("VPREB3", "CDH5", "CLDN5", "ECSCR", "TIE1", "MIA", "BANK1", "PAX5", "CD22")

## For each cell type, generate model using stepAIC with the high VIF genes removed
all_models<- lapply(cell_types, function(cell_type){
  train<- train.test_byCellType[[cell_type]]$train
  train<- train[,!colnames(train) %in% highVIF]
  train$Dependent<- train[[cell_type]]; train[[cell_type]]<-NULL
  model<- lm(Dependent ~ ., train)
  stepAIC_model<- stepAIC(model, trace=FALSE)
  return(stepAIC_model)
})
names(all_models)<-cell_types
```

## Plot genes included in model for each cell type
```{r}
## Get final gene input into each model
gene_input<- lapply(names(all_models), function(x){
  model_summary<- summary(all_models[[x]])
  pvalues<- model_summary$coefficients[,"Pr(>|t|)"]
  pvalues<- as.data.frame(pvalues[!grepl("Intercept", names(pvalues))])
  pvalues$gene<- rownames(pvalues); pvalues$cellType<-x; rownames(pvalues)<-NULL
  return(pvalues)
})
gene_input<- do.call("rbind", gene_input)

## Plot each gene's contribution to model (pvalue)
ggplot(gene_input, aes(x=gene, y=pvalues, color=cellType))+
  geom_point()+
  theme_bw()+
  facet_grid(cols = vars(cellType), space="free_x", scale="free_x")+
  geom_hline(yintercept = 0.05, linetype="dashed")+
  theme(axis.text.x = element_text(angle = 45, hjust=1, face="bold", color="black"),
        legend.position = "none")+
  ylab("pvalue") + xlab("gene input into model")
ggsave(paste0(figure_dir, "Figure3_LR_genesInputIntoModel.png"), width=8, height=3)
```


## For each cell type, use the model above to predict cell type proportions in testing data
```{r}
## For each cell type, apply model above to predict cell type proportion for testing data
all_predict<- lapply(cell_types, function(cell_type){
  test<- train.test_byCellType[[cell_type]]$test
  model<- all_models[[cell_type]]
  predict<- predict(model, test)
  test$prediction<- predict
  return(test)
})
names(all_predict)<- cell_types

## Plot actual vs predicted
all_plots<- lapply(cell_types, function(cell_type){
  df<- all_predict[[cell_type]]
  df$Actual<- df[[cell_type]]
  
  ## Get r2/pvalue for actual vs predicted
  model<- summary(lm(Actual ~ prediction, df))
  r2<- round(model$r.squared, 2)
  pvalue<- formatC(model$coefficients["prediction", "Pr(>|t|)"],format="e", digits=2)
  ggplot(df, aes(x=Actual, y=prediction))+
    geom_point()+
    theme_bw()+
    ggtitle(cell_type,subtitle =  paste0("R2 = ", r2, ", pvalue = ", pvalue))+
    ylab(paste0("Predicted proportion\nof ", cell_type))+
    xlab(paste0("Actual proportion\nof ", cell_type))+
    geom_smooth(method="lm", se=FALSE, color="red", linetype="dashed")+
    theme(axis.title = element_text(color="black", face="bold"),
          axis.text = element_text(color="black", face="bold"),
          title  = element_text(color="black", face="bold"))
})
cowplot::plot_grid(plotlist=all_plots)
ggsave(paste0(figure_dir, "Figure4_LinearRegression_PredictCellProportion_testing.png"), width=10, height=6)
```

## Apply linear regression models to bulk data to predict cell type proportions
```{r}
## Only select models that highly significantly predict (p<0.01) cell type in pseudobulked data
sig_cellTypes<- c("B.cells", "CAFs", "Endothelialcells", "Macrophages", "melanoma")
sig_models<- all_models[names(all_models) %in% sig_cellTypes]

## Transpose bulk data
bulk_t<- as.data.frame(t(bulk_cm))
colnames(bulk_t)<- gsub("-", ".", colnames(bulk_t))

## For each model, predict cell type proportion
all_bulk_predict<- lapply(names(sig_models), function(x){
  model_use<- sig_models[[x]]
  bulk_predict<- predict(model_use, bulk_t)
  bulk_predict<- as.data.frame(bulk_predict)
  colnames(bulk_predict)<-paste0("predicted_", x)
  return(bulk_predict)
})
all_bulk_predict<- do.call("cbind", all_bulk_predict)


## Plot boxplot of results
ggplot(tmp, aes(x=CellType, y=Proportion, fill=CellType))+
  geom_boxplot()+
  theme_bw()+
  theme(legend.position = "none")+
  ylab("Predicted cell type composition")+xlab("")+
  ggtitle("Predicted cell type composition in bulk data\nbased on linear regression modeling")
ggsave(paste0(figure_dir, "Figure5_LR_boxPlot_predictCompInBulk.png"), width = 5, height=5)

write.csv(all_bulk_predict, file=paste0(analysis_dir, "linearRegression_bulk_predicted_cellProp.csv"))

```

