---
title: "R Notebook"
output: html_document
---

## Libraries
```{r}
library(dplyr)
library(MASS)
library(ggplot2)
library(mlbench)
library(caret)
library(e1071)
library(lime)
library(randomForest)
```

```{r}
base_dir = "/Users/jlabelle/Dropbox (Partners HealthCare)/"
working_dir<- paste0(base_dir, "Personal/STAT_E109/Assignments/Project/")

## Set output directories
data_dir<- "analysis/predictCellTypeComp/"
analysis_dir<- "analysis/predictTumorGrade/"
figure_dir<- "figures/"
```

## Read in predicted cell type scores for bulk data (based on linear regression model and random forest) and clinical data
```{r}
predicted_scores<- read.csv(paste0(data_dir, "linearRegression_bulk_predicted_cellProp.csv"), row.names = "X")
clinical_all<- read.delim("data/clinical.cases_selection.2022-04-10/clinical.tsv")


clinical<- clinical_all
```

## Reformat clinical data
```{r}
## Rename sample ids to match
clinical$case_id <- gsub("-", ".", clinical$case_id)
sum(clinical$case_id %in% rownames(predicted_scores))

## Not all cases have clinical info- subset to patients with both
predicted_scores<- predicted_scores[rownames(predicted_scores) %in% clinical$case_id,]
clinical<- clinical[clinical$case_id %in% rownames(predicted_scores),]

## Remove non-unique clinical info- from multiple samples. Clinical info is the same.
clinical<- clinical[!duplicated(clinical$case_id),]

## Subset clinical to patient outcome variables of interest + potential variables to control for
clinical<-clinical[,c("case_id", "gender", "age_at_index","ajcc_pathologic_stage")]

## Stage includes substages- reduce to simplify and increase number of cases per stage
clinical$ajcc_pathologic_stage<- gsub("A", "", gsub("B", "", gsub("C", "", clinical$ajcc_pathologic_stage)))
```

## Merge clinical and predicted scores
```{r}
predicted_scores$case_id<- rownames(predicted_scores)
clinical_scores<- merge(clinical, predicted_scores, by="case_id")

## Remove cases without info on stage
clinical_scores<- clinical_scores[!clinical_scores$ajcc_pathologic_stage %in% c("'--", "Not Reported"),]
clinical_scores<- clinical_scores[,!colnames(clinical_scores) %in% c("case_id")]

## Convert age to numeric
clinical_scores$age_at_index<- as.numeric(clinical_scores$age_at_index)

## Remove stage 0
input<- clinical_scores[!clinical_scores$ajcc_pathologic_stage %in% c("Stage 0"), ]

## Set outcome variable to factor
input$ajcc_pathologic_stage<- factor(input$ajcc_pathologic_stage)
```


## Generate models using bagging, random forest, and boosting. Assess accuracy of each model
```{r}
## Split into training/testing
set.seed(5)
ind <- sample(2, nrow(input), replace = T, prob = c(0.7, 0.3))
train <- input[ind == 1,]
test <- input[ind == 2,]

## Df to save all results in
cmat_res<- data.frame(Model="NA", Accuracy="NA", AccuracyNull="NA", pvalue="NA")

## Training settings
set.seed(1234) 
cvcontrol <- trainControl(method="repeatedcv", 
                          number = 5,
                          repeats = 1,
                          allowParallel=TRUE)

# Bagging
set.seed(1234) 
  ## generate model
  bag <- train(ajcc_pathologic_stage ~ predicted_B.cells + predicted_Macrophages + predicted_Endothelialcells + predicted_CAFs + predicted_melanoma+ age_at_index +gender, 
               data=train,
               method="treebag",
               trControl=cvcontrol,
               importance=TRUE)
  
  ## Plot important features
  png(paste0(figure_dir, "Figure6A_lrBased_baggingImportantFeatures.png"), width=500, height=300)
  plot(varImp(bag), main="Bagging")
  dev.off()

  ## Accuracy
  p <- predict(bag, test, type = 'raw')
  cmat<- confusionMatrix(p, test$ajcc_pathologic_stage)
  cmat_res<-rbind(cmat_res, data.frame(Model="Bagging", 
                                       Accuracy=round(cmat$overall["Accuracy"],3),
                                       AccuracyNull=round(cmat$overall["AccuracyNull"],3),
                                       pvalue=round(cmat$overall["AccuracyPValue"], 6)))

# RF
set.seed(1234)
  ## Generate model
  forest <- train(ajcc_pathologic_stage ~ predicted_B.cells + predicted_Macrophages + predicted_Endothelialcells + predicted_CAFs + predicted_melanoma+ age_at_index +gender , 
                  data=train,
                  method="rf",
                  trControl=cvcontrol,
                  importance=TRUE)
  
  ## Plot important features
  png(paste0(figure_dir, "Figure6B_lrBased_forestImportantFeatures.png"), width=500, height=300)
  plot(varImp(forest), main="Forest")
  dev.off()
  
  ## Accuracy
  p <- predict(forest, test, type = 'raw')
  cmat<- confusionMatrix(p, test$ajcc_pathologic_stage)
  cmat_res<-rbind(cmat_res, data.frame(Model="Forest", 
                                       Accuracy=round(cmat$overall["Accuracy"],3),
                                       AccuracyNull=round(cmat$overall["AccuracyNull"],3),
                                       pvalue=round(cmat$overall["AccuracyPValue"], 6)))

# Boosting
set.seed(1234)

  ## Generate model
  boo <- train(ajcc_pathologic_stage ~ predicted_B.cells + predicted_Macrophages + predicted_Endothelialcells + predicted_CAFs + predicted_melanoma+ age_at_index +gender, 
               data=train,
               method="xgbTree",   
               trControl=cvcontrol,
               tuneGrid = expand.grid(nrounds = 500,
                                      max_depth = 4,
                                      eta = 0.28,
                                      gamma = 1.8,
                                      colsample_bytree = 1,
                                      min_child_weight = 1,
                                      subsample = 1))
  
  ## Plot important features
  png(paste0(figure_dir, "Figure6C_lrBased_boostingImportantFeatures.png"), width=500, height=300)
  plot(varImp(boo), main="Boosting")
  dev.off()
  
  ##  Accuracy
  p <- predict(forest, test, type = 'raw')
  cmat<- confusionMatrix(p, test$ajcc_pathologic_stage)
  cmat_res<-rbind(cmat_res, data.frame(Model="Boosting", 
                                       Accuracy=round(cmat$overall["Accuracy"],3),
                                       AccuracyNull=round(cmat$overall["AccuracyNull"],3),
                                       pvalue=round(cmat$overall["AccuracyPValue"], 6)))

```

## Does random forest model predict tumor grade better than age + gender alone?
```{r}
set.seed(1234) 
cvcontrol <- trainControl(method="repeatedcv", 
                          number = 5,
                          repeats = 1,
                          allowParallel=TRUE)
forest_ageOnly <- train(ajcc_pathologic_stage ~  age_at_index +gender, 
             data=train,
             method="rf",
             trControl=cvcontrol,
             importance=TRUE)
p <- predict(forest_ageOnly, test, type = 'raw')
confusionMatrix(p, test$ajcc_pathologic_stage)
## Accuracy: 0.3659, lower than no information rate of 0.39, pvalue=0.432
```

## Save table with results from all 3 models
```{r}
df<-cmat_res[cmat_res$Model!="NA",]
rownames(df)<-NULL

png(paste0(figure_dir, "Table3_decisionTreeModel_results.png"), height=100, width=500)
p<-tableGrob(df)
grid.arrange(p)
dev.off()
```

## For random forest: plot predictions
```{r}
p <- predict(forest, test)
cmat<- as.data.frame(confusionMatrix(p, test$ajcc_pathologic_stage)$table)

## Plot number of samples assigned vs actual tumor grade
p1<-ggplot(cmat, aes(x=Reference, y=Freq, fill=Prediction))+
  geom_bar(stat="identity")+
  xlab("Actual tumor stage")+
  ylab("Predicted tumor stage (n samples)")+
  theme_classic()+
  theme(axis.text.x = element_text(angle=45, hjust=1))+
  ggtitle("Random Forest model predictions:\nnumber of samples")

## Conver to proportion
cmat<- cmat %>% group_by(Reference) %>% mutate(perFreq=Freq/sum(Freq)) %>% as.data.frame()
p2<-ggplot(cmat, aes(x=Reference, y=perFreq, fill=Prediction))+
  geom_bar(stat="identity")+
  xlab("Actual tumor stage")+
  ylab("Predicted tumor stage (% samples)")+
  theme_classic()+
  theme(axis.text.x = element_text(angle=45, hjust=1))+
  ggtitle("Random Forest model predictions:\npercentage of samples")

cowplot::plot_grid(plotlist=list(p1, p2))


ggsave(paste0(figure_dir, "Figure7_lrBased_RandomForest_predict.vs.actual_testing.png") ,width=7, height=3)
```